---
title: "Fake data"
output: html_document
---

```{r setup}
library(tidyverse)
library(wakefield)
library(scales)
```

```{r}
set.seed(1234)
ds <- data_frame(income = rnorm(2000, 5000, 1000),
                 health = rnorm(2000, 3, 1)) %>% 
  mutate(health = rescale(health, to = c(1, 5))) %>% 
  mutate(health = round(health, digits = 0)) %>% 
  mutate(income = rescale(income, c(15000, 400000))) %>% 
  mutate(income_adjusted = income * health)
  

ggplot(data = ds, mapping = aes(x = income)) +
  geom_histogram(color = "white")
```


```{r}
ds1 <- r_data_frame(
  n = 2000,
  id,
  age = age(5:18),
  gender_inclusive,
  sex_inclusive
) %>% 
  mutate(income = rnorm(2000, 5000, 1000),
         health = rnorm(2000, 3, 1)) %>% 
  mutate(treatment = sample(c("Treatment", "Control"), 2000, replace = TRUE)) %>% 
  mutate(bed_net = sample(c("Bed Net", "No bed net"), 2000, replace = TRUE)) %>% 
  mutate(income = ifelse(bed_net == TRUE, income + 10000, income))
  
```







```{r}
# Make all the random draws consistent
set.seed(1234)

# The r_data_frame() function lets you generate random data. You just have to
# feed it the name of a function that generates variables. You can see all the
# possible variable-generating functions by running variables() in the R
# console:
variables()

# Here we generate a small data frame. Each of the functions that are listed in variables() has arguments. See age, for instance---we can specify a range of ages to draw from
small_data <- r_data_frame(
  n = 30,
  age(5:18),  # Young kids
  income(digits = 5),  # Rich kids
  zip_code,
  race,
  gender_inclusive
)

# Look at the first few rows
head(small_data)

# You can make more complicated data too, like adding normally-distributed
# income, or assigning people to treatment and control groups
treatment <- r_data_frame(
  n = 500,
  race,
  age(5:18),
  income = rnorm(mean = 100000, sd = 15000)  # Normal distribution centered at 100000
) %>% 
  mutate(treatment = "Yes")

control <- r_data_frame(
  n = 500,
  race,
  age(5:18),
  income = rnorm(mean = 50000, sd = 15000)  # Normal distribution centered at 50000
) %>% 
  mutate(treatment = "No")

# We can combine these two datasets into one with bind_rows(), which essentially
# stacks the rows of one on top of the rows of the other:
big_data_set <- bind_rows(treatment, control)

# For fun, we can check the difference in income for people in the treatment and control groups
ggplot(big_data_set, aes(x = income, fill = treatment)) + 
  geom_density()
```

Whoa! Look at that! The imaginary program boosted incomes substantially! :)

In problem set 5 you worked with fake data about a hypothetical math camp. Here's the code I used to generate that data. The only odd thing here is the `rtruncnorm()` function, which generates data from a truncated normal distribution. This makes it so you can put limits on numbers—if you want a random distribution of GPAs centered at 3.5, you don't want to accidentally create GPAs of 4.3 or whatever. The `a` and `b` arguments let you set a minimum and a maximum number.

```{r eval=FALSE}
# Make all the random draws consistent
set.seed(1234)

# Treatment group with higher post-treatment GPA
treatment <- r_data_frame(n = 794, 
                          id, 
                          race, 
                          age(x = 20:30), 
                          gender_inclusive) %>% 
  mutate(undergrad_gpa = round(rtruncnorm(n(), a = 1.0, b = 4.0, 
                                          mean = 2.5, sd = .5), 2),
         math_camp = TRUE,
         gre_verbal = round(rtruncnorm(n(), a = 130, b = 170,
                                       mean = 145, sd = 15), 0),
         gre_quant = round(rtruncnorm(n(), a = 130, b = 170,
                                      mean = 110, sd = 15), 0),
         gre_total = gre_verbal + gre_quant,
         graduate_gpa = round(rtruncnorm(n(), a = 1.0, b = 4.0, 
                                         mean = 3.3, sd = .5), 2))

# Control group with slightly higher post-treatment GPA
control <- r_data_frame(n = 787, 
                        id, 
                        race, 
                        age(x = 20:30), 
                        gender_inclusive) %>% 
  mutate(undergrad_gpa = round(rtruncnorm(n(), a = 1.0, b = 4.0, 
                                          mean = 2.5, sd = .5), 2),
         gre_verbal = round(rtruncnorm(n(), a = 130, b = 170,
                                       mean = 145, sd = 15), 0),
         gre_quant = round(rtruncnorm(n(), a = 130, b = 170,
                                      mean = 110, sd = 15), 0),
         math_camp = FALSE,
         gre_total = gre_verbal + gre_quant,
         graduate_gpa = round(rtruncnorm(n(), a = 1.0, b = 4.0, 
                                         mean = 2.9, sd = .5), 2))

population <- r_data_frame(n = 1986, 
                           id, 
                           race, 
                           age(x = 20:30), 
                           gender_inclusive) %>% 
  mutate(undergrad_gpa = round(rtruncnorm(n(), a = 1.0, b = 4.0, 
                                          mean = 2.5, sd = 1.5), 2),
         gre_verbal = round(rtruncnorm(n(), a = 130, b = 170,
                                       mean = 145, sd = 15), 0),
         gre_quant = round(rtruncnorm(n(), a = 130, b = 170,
                                      mean = 160, sd = 30), 0),
         math_camp = FALSE,
         gre_total = gre_verbal + gre_quant,
         graduate_gpa = round(rtruncnorm(n(), a = 1.0, b = 4.0, 
                                         mean = 3.6, sd = 1), 2))

# Combine them all
experiment <- bind_rows(treatment, control) %>% 
  sample_frac(1)  # Shuffle the dataset for kicks and giggles

everyone <- bind_rows(experiment, population) %>% 
  sample_frac(1)

# Save these as CSV files
write_csv(experiment, "math_camp_experiment.csv")
write_csv(everyone, "math_camp_everyone.csv")
```

