---
title: "Playing with R Markdown"
---

# Literate programming

R Markdown lets you combine text with code so that you no longer have to copy/paste numbers, tables, and graphs from one program to another! R Markdown is really just the combination of two different things: Markdown and R.

## Markdown

Markdown is a super simple syntax you can use to style your text. If you want something italicized, *add asterisks around it*; if you want something bolded, **add two asterisks around it.** You can

- make
- bulleted
- lists

And also

1. make
2. numbered
3. lists

You can make headings and subheadings with #s. 1 # = a 1st level heading, 2 #s = a 2nd level heading, etc. 

[Here's a quick introduction to a bunch of other Markdown features.](https://guides.github.com/features/mastering-markdown/)


## R within Markdown

The magic of R Markdown is that it lets you mix R code and output inside text. Most often you'll put R code in what are called "chunks", which look like this:

```{r}
2 + 2
```

When you compile this document, a package named "knitr" will run each chunk and include its output in the document (because of this, compiling an R Markdown file is generally called "knitting").

Here's an example regression model:

```{r}
mpg_model <- lm(mpg ~ wt, data = mtcars)
summary(mpg_model)
```

Neat! It's kind of ugly when you knit, but later I'll show you how to make this look pretty.

You can also include graphics:

```{r}
library(ggplot2)

ggplot(data = mtcars, 
       mapping = aes(x = wt, y = mpg)) +
  geom_point(mapping = aes(color = as.factor(cyl))) +
  geom_smooth(method = "lm")
```

If you want to run R code within a paragraph of text instead of in a chunk, you can do that too. Like, 2 + 2 is equal to `r 2 + 2`, or pi rounded to the first 5 digits is `r round(pi, 5)`.

## Output formats

With the magic of R Markdown, you can knit your document to a bunch of different formats:

- HTML
- Word
- PDF (requires some extra installation)
