---
title: "DAGs with R"
output: html_document
---

The **ggdag** and **dagitty** packages make it easy to plot and work with the logic of DAGs.

```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggdag)
library(dagitty)
```

# Education and earnings

Here's the education and earnings DAG we did in daggity:

```{r}
educ_earn_dag <- dagify(Earn ~ Edu + Year + Bkgd + Loc + JobCx,
                        Edu ~ Loc + Bkgd + Year + Req,
                        JobCx ~ Edu,
                        Loc ~ U1,
                        Bkgd ~ U1,
                        outcome = "Earn",
                        exposure = "Edu",
                        latent = "U1")
```

Plot it:

```{r}
ggdag(educ_earn_dag)
```

Plot it with a nicer theme:

```{r}
ggdag(educ_earn_dag) + 
  theme_dag()
```

What are all the possible paths between treament and outcome?

```{r}
paths(educ_earn_dag)
```

We can plot all the paths:

```{r}
ggdag_paths(educ_earn_dag, shadow = TRUE) +
  theme_dag()
```

We can figure out the adjustment set automatically too:

```{r}
ggdag_adjustment_set(educ_earn_dag, shadow = TRUE) +
  theme_dag()
```


# Wine and lifespan

Here's a DAG showing the relationship between wine consumption and lifespan:

```{r}
wine_lifespan_dag <- dagify(lifespan ~ wine + drugs + income + health,
                            drugs ~ wine,
                            wine ~ income + health,
                            health ~ something,
                            income ~ something,
                            latent = "something",
                            exposure = "wine",
                            outcome = "lifespan",
                            labels = c("lifespan" = "Lifespan",
                                       "wine" = "Drinking wine",
                                       "drugs" = "Drug use",
                                       "income" = "Income",
                                       "health" = "Health",
                                       "something" = "Something else"))

# With fancier labels
ggdag(wine_lifespan_dag, text = FALSE, use_labels = "label") +
  theme_dag()
```

Possible paths:

```{r}
paths(wine_lifespan_dag)
```

Adjustment set:

```{r}
ggdag_adjustment_set(wine_lifespan_dag, text = FALSE, use_labels = "label", shadow = TRUE) +
  theme_dag()
```

