---
title: "Playground"
output: html_document
---

Messages: don't worry about them!

```{r}
library(tidyverse)
library(broom)
library(gapminder)
```

Warnings: Okay-ish, but follow what they say:

```{r}
ggplot(gapminder, aes(x = lifeExp, fill = continent)) + 
  # geom_histogram(binwidth = 1)
  geom_density(alpha = 0.5)# +
  facet_wrap(~ continent)
```

Errors: Something is wrong, and R will try to tell you what it is.

```{r eval=FALSE}
ggplot(gampinder, aes(x = lifeExp)) + 
  geom_histogram()
```

# Extracting things from models

```{r}
gapminder_2007 <- gapminder %>% 
  filter(year == 2007)

example_model <- lm(lifeExp ~ log(gdpPercap) + continent, data = gapminder_2007)

example_model_df <- tidy(example_model)
```

The effect of blah is 12.9 in Oceania. 

One way is to use `$`, but this gets the whole column:

```{r}
example_model_df$estimate

example_model_df$estimate

```

Another way is to filter the data frame first:

```{r}
example_model_df %>% 
  filter(term == "log(gdpPercap)")
```

This gives you all the columns for one specific row. You can get just the value you want by adding `pull()`, which pulls that value out:

```{r}
example_model_df %>% 
  filter(term == "log(gdpPercap)") %>% 
  pull(p.value)
```

If you know the row number you care about, you can use `slice()` instead of `filter()` to just select that one row:

```{r}
oceanian_estimate <- example_model_df %>% 
  slice(6) %>% 
  pull(estimate)
```

Oceania's estimate is `r round(oceanian_estimate, 2)`. The p value is QWERQWE and the t-statistic is QWERQW ;l

This works for any data frame. Here's how to get the $R^2$ value out, for example:



```{r}
model_details <- glance(example_model)

# The dollar sign way, since there's only one row:
model_details$adj.r.squared

# The pull way:
model_details %>% pull(adj.r.squared)
```

You can then use this in your text and say that the model has fairly good fit ($R^2$ = `r model_details %>% pull(adj.r.squared) %>% round(2)`).

Or you can store the value as a variable to make typing easier:

```{r}
model_r2 <- model_details %>% pull(adj.r.squared)
```

The model's $R^2$ is `r round(model_r2, 2)`


# Math

Markdown lets you use fancy math typesetting notation called LaTeX (see https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html for some examples of how to use it). 

There are two ways to use LaTeX math: inline within a paragraph, and as a standalone paragraph. Inline, math is surrounded by dollar signs: $\beta_1$, or $y = mx + b$, etc. When putting it in a standalong paragraph, it's surrounded by two dollar signs:

$$
y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon
$$

Because the dollar sign is used for mathy things, if you want to type an actual dollar sign in Markdown, you'll need to put a backslash in front of it: the cost of a gallon of milk is \$2.98. 


