---
title: "World happiness"
output: html_document
---


```{r load-libraries-data, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)

world_happiness <- read_csv("data/world_happiness.csv")
```

# Correlation

What is the correlation between happiness and life expectancy? Do they move together?

```{r find-correlation}
world_happiness %>% 
  summarize(happy_life_correlation = cor(happiness_score, life_expectancy,
                                         use = "complete.obs"))
```

What does this look like?

```{r}
ggplot(world_happiness, mapping = aes(x = life_expectancy, y = happiness_score)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

# Simple model

We know that happiness and life expectancy are correlated, but how are they associated? How do they move together?

```{r}
model1 <- lm(happiness_score ~ life_expectancy, data = world_happiness)
```

There are different ways of seeing the model results. You can use `summary()` to see all the details:

```{r}
summary(model1)
```

But this has a lot of extra information and it's ugly and hard to read in a document. It'd be nicer if this was in a table. We can make a table of the model coefficients with the `tidy()` function from the **broom** package:

```{r}
tidy(model1)
```

If you want confidence intervals for the estimates (you generally do), use `tidy(model1, conf.int = TRUE)`:

```{r}
tidy(model1, conf.int = TRUE)
```


And we can get the $R^2$ value and other diagnostic numbers with `glance()`:

```{r}
glance(model1)
```

## Simple model interpretation

Given these results, we can write the model as an equation:

$$
\hat{\text{happiness}} = -2.22 + 0.105 \times \text{life expectancy} + \epsilon
$$

The intercept of -2.22 means that this line crosses the y-axis at -2.22. In a hypothetical country with a life expectancy of 0, the happiness score would be -2. 

What is important, though, is the value of the β1

coefficient, or the slope of life expectancy. This tells us that for every unit increase in life expectancy, there’s an associated increase in happiness of 0.105. If life expectancy increase by 10 years, we could predict an associated increase in happiness of 1.05 points, on average. Neat!

We can use the following template to interpret this coefficient:

> A one unit increase in life expectancy is associated with a 0.105 point increase in happiness, on average.

Don't succumb to the temptation to tell a causal story, though. Life expectancy doesn't necessarily cause happiness. The reverse story is just as plausible (as life expectancy increases, happiness will increases). There's no mathematical or statistical way to test for causation though—it's all a question of theory and philosophy.

# Bigger model

We can control for other factors as well by include more variables in our model:

```{r}
model2 <- 
  lm(happiness_score ~ life_expectancy + school_enrollment + region, 
     data = world_happiness)
```

Let's check the coefficients now:

```{r}
tidy(model2, conf.int = TRUE)
```

```{r}
glance(model2)
```

