---
title: "DAGs with R"
output: html_document
---

The **ggdag** and **dagitty** packages make it easy to plot and work with the logic of DAGs.

```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggdag)
library(dagitty)
```

# Education and earnings

Here's the education and earnings DAG we did in daggity:

```{r}
educ_earn_dag <- dagify(Earn ~ Edu + Year + Bkgd + Loc + JobCx,
                        Edu ~ Loc + Bkgd + Year + Req,
                        JobCx ~ Edu,
                        Loc ~ U1,
                        Bkgd ~ U1,
                        outcome = "Earn",
                        exposure = "Edu",
                        latent = "U1")
```

Plot it:

```{r}
ggdag(educ_earn_dag)
```

Plot it with a nicer theme:

```{r}
ggdag(educ_earn_dag) + 
  theme_dag()
```

What are all the possible paths between treament and outcome?

```{r}
paths(educ_earn_dag)
```

We can plot all the paths:

```{r}
ggdag_paths(educ_earn_dag, shadow = TRUE) +
  theme_dag()
```

We can figure out the adjustment set automatically too:

```{r}
ggdag_adjustment_set(educ_earn_dag, shadow = TRUE) +
  theme_dag()
```


# Wine and lifespan

Here's a DAG showing the relationship between wine consumption and lifespan:

```{r}
wine_lifespan_dag <- dagify(lifespan ~ wine + drugs + income + health,
                            drugs ~ wine,
                            wine ~ income + health,
                            health ~ something,
                            income ~ something,
                            latent = "something",
                            exposure = "wine",
                            outcome = "lifespan",
                            labels = c("lifespan" = "Lifespan",
                                       "wine" = "Drinking wine",
                                       "drugs" = "Drug use",
                                       "income" = "Income",
                                       "health" = "Health",
                                       "something" = "Something else"))

# With fancier labels
ggdag(wine_lifespan_dag, text = FALSE, use_labels = "label") +
  theme_dag()
```

Possible paths:

```{r}
paths(wine_lifespan_dag)
```

Adjustment set:

```{r}
ggdag_adjustment_set(wine_lifespan_dag, text = FALSE, use_labels = "label", shadow = TRUE) +
  theme_dag()
```

# Adjusting without regression

In 1986 Robert LaLonde published a report that studied the effect of a job training program on earnings in 1978 (`re78`. His study used an experiment to assign people to a training program or not, and he also collected data on people who voluntarily joined, which includes their earnings in 1975 (`re75`), prior to the training program.

```{r}
# This package contains the data we want
library(wooldridge)

# People who participated in training by random assignment
data("jtrain2")

# People who participated in training by choice
data("jtrain3")
```

What were the hourly wages of people in the experiment, both in and out of the training program?

```{r}
jtrain2 %>% 
  group_by(train) %>% 
  summarize(avg_wage = mean(re78))
```

Fairly similar. Without any inferential tools, it looks like the training people earned almost \$2/hour more.

What about the general public? People who voluntarily received training?

```{r}
jtrain3 %>% 
  group_by(train) %>% 
  summarize(avg_wage = mean(re78))
```

Whoa! Those without the training are waaay richer. The training program is making people poorer?

There are lots of confounders though. People who need the training want to get it it but are likely to have lower wages already (since they need the training).

```{r}
# Possible DAG
training_dag <- dagify(wage ~ train + U + need_tr,
                       train ~ U + need_tr,
                       exposure = "train",
                       outcome = "wage")

ggdag(training_dag)
```

Let's adjust for needing training. These two datasets are looking at different groups of people. Look at their incomes before the training program was offered, `re75`:

```{r}
# Experimental people
ggplot(jtrain2, aes(x = re75)) +
  geom_histogram(binwidth = 2)
```

```{r}
# Non-experimental people
ggplot(jtrain3, aes(x = re75)) +
  geom_histogram(binwidth = 2)
```

We can look at the distribution of these incomes:

```{r}
summary(jtrain2$re75)  # Experiment
summary(jtrain3$re75)  # Non experiment
```

Look at the 3rd quartile, or 75th percentile. 75% of the people in the experiment had an hourly wage of \$1.20 or less, while 75% of the people in the general public had a wage of \$25 or less!

What if we limit the nonexperimental sample to just people who earned \$1.20 or less?

```{r}
jtrain3 %>% 
  filter(re75 <= 1.20) %>% 
  group_by(train) %>% 
  summarize(avg_wage = mean(re78))
```

This isn't perfect. The \$1.20 cutoff was arbitrary, and we still have the `train <- U -> wage` backdoor to close, but it's an improvement! Choosing a sample is a form of controlling.

# Colliders and DAGs

```{r}
tech_dag <- dagify(social ~ prog,
                   hired ~ prog + social,
                   exposure = "prog",
                   outcome = "social",
                   labels = c(social = "Social skills", prog = "Programming skills",
                              hired = "Hired by a tech company"))

ggdag(tech_dag, use_labels = "label") +
  theme_dag()
```

```{r}
ggdag_collider(tech_dag)
ggdag_adjustment_set(tech_dag)
```

