---
title: "Effect of Katrina evacuation on employment"
output: html_document
---

```{r load-libraries-data, warning=FALSE, message=FALSE}
library(tidyverse)
library(haven)  # For reading files from Stata, SPSS, etc.
library(estimatr)  # For robust regression and standard errors

katrina <- read_stata("data/katrina.dta") %>% 
  # Clean up some of the columns so they have labels instead of 0/1, 1/2/3/4/5, etc.
  mutate(health_cat = factor(health, labels = c("Excellent", "Very good", "Good", "Fair", "Poor"),
                             ordered = TRUE),
         everevac_bin = factor(everevac, labels = c("Not evacuated", "Evacuated")),
         year_factor = factor(year, ordered = TRUE),
         storm = factor(ifelse(year == 2005, "Before", "After"), levels = c("Before", "After")))
```


# Exploratory analysis

We're interested in the effect of Katrina evacuation on four different outcomes: time spent employed, earnings, unemployment income, and health. 

We first want to see what the effect of the storm was on those who evacuated. What are the averages and distributions for these different outcomes?

Here we're only looking at those who evacuated, and only in 2006 after they evacuated (this is like an experiment with no control group and no pretest).

```{r}
# Only look at people in 2006 who had evacuated
katrina_2006_evac <- katrina %>% 
  filter(year == 2006, everevac_bin == "Evacuated")
```

## Weeks worked

```{r}
katrina_2006_evac %>% 
  summarize(avg_wks_worked = mean(wkswork))
```


```{r}
ggplot(katrina_2006_evac, aes(x = wkswork)) +
  geom_density(fill = "blue")
```


## Earnings

```{r}
# Get the average for earnings
```

```{r}
# Plot the distrubtion for earnings
```


## Unemployment income

```{r}
# Get the average for unempinc
```

```{r}
# Plot the distrubtion for unempinc
```


## Health

We can't calculate the average health, since it's categorical. But we can look at the distribution:

```{r}
# Make a smaller summary dataframe first
health_plot <- katrina_2006_evac %>% 
  group_by(health_cat) %>% 
  summarize(n = n()) %>% 
  # Get the percentage for each category
  ungroup() %>% 
  mutate(prop = n / sum(n))

ggplot(health_plot, aes(x = health_cat, y = prop)) + 
  geom_col()
```


# Before/after comparison for evacuees

Next, calculate the effect of evacuation for just those who evacuated (i.e. ignore those who didn't evacuate; there's no control group). 

Check the difference in means visually and with a t-test. Run a regression with `storm` as the explanatory variable, then with `storm` and controls for age, race, sex, and education (`age + sex + black + hsgrad + someco + ba + postgrad`). 

Without looking at a control group, did the storm have any effect on evacuees in weeks worked, earnings, and unemployment income? (don't worry about health for now). Does adding controls change anything?

```{r}
# Only look at evacuees
katrina_evac <- katrina %>% 
  filter(everevac_bin == "Evacuated")
```

## Weeks worked

```{r}
# Difference in means
t.test(wkswork ~ storm, data = katrina_evac)

model_work_evac <- lm(wkswork ~ storm, data = katrina_evac)

model_work_evac_full <- lm(wkswork ~ storm + age + sex + black + 
                             hsgrad + someco + ba + postgrad,
                           data = katrina_evac)
tidy(model_work_evac)
tidy(model_work_evac_full)
```

```{r}
# Calculate the average weeks worked + standard error
wks_worked_plot <- katrina_evac %>% 
  group_by(storm) %>% 
  summarize(avg = mean(wkswork),
            sd = sd(wkswork),
            n = n(),
            se = sd / sqrt(n))

ggplot(wks_worked_plot, aes(x = storm, y = avg, color = storm)) +
  geom_errorbar(aes(ymin = avg - 1.96 * se,
                    ymax = avg + 1.96 * se), 
                size = 1, width = 0.1) +
  geom_point() +
  geom_line(aes(group = 1), color = "grey50") +
  guides(color = FALSE)  # Turn off legend
```

## Earnings

```{r}
# t-test and models for earnings
```

```{r}
# Plot for earnings
```

## Unemployment income

```{r}
# t-test and models for unempinc
```

```{r}
# Plot for unempinc
```

# Evacuee / non-evacuee comparison in 2006

Next, calculate the effect of evacuation for just those who evacuated (i.e. ignore 2005; there's no pretest). 

Check the difference in means visually and with a t-test. Run a regression with `everevac_bin` as the explanatory variable, then with `everevac_bin` and controls for age, race, sex, and education (`age + sex + black + hsgrad + someco + ba + postgrad`). 

Without looking at pre-storm levels, did the storm have any effect on evacuees in weeks worked, earnings, and unemployment income? (don't worry about health for now). Does adding controls change anything?

```{r}
# Only look at evacuees
katrina_2006 <- katrina %>% 
  filter(year == 2006)
```

## Weeks worked

```{r}
# Difference in means
t.test(wkswork ~ everevac_bin, data = katrina_2006)

model_work_2006 <- lm(wkswork ~ everevac_bin, data = katrina_2006)

model_work_2006_full <- lm(wkswork ~ everevac_bin + age + sex + black + 
                             hsgrad + someco + ba + postgrad,
                           data = katrina_2006)
tidy(model_work_2006)
tidy(model_work_2006_full)
```

```{r}
# Calculate the average weeks worked + standard error
wks_worked_plot <- katrina_2006 %>% 
  group_by(everevac_bin) %>% 
  summarize(avg = mean(wkswork),
            sd = sd(wkswork),
            n = n(),
            se = sd / sqrt(n))

ggplot(wks_worked_plot, aes(x = everevac_bin, y = avg, color = everevac_bin)) +
  geom_errorbar(aes(ymin = avg - 1.96 * se,
                    ymax = avg + 1.96 * se), 
                size = 1, width = 0.1) +
  geom_point() +
  geom_line(aes(group = 1), color = "grey50") +
  guides(color = FALSE)  # Turn off legend
```

## Earnings

```{r}
# t-test and models for earnings
```

```{r}
# Plot for earnings
```

## Unemployment income

```{r}
# t-test and models for unempinc
```

```{r}
# Plot for unempinc
```


# Difference-in-differences

Finally (!!) we can see if the evacuation itself had an effect, account for treatment/control and before/after the storm. Interpret these coefficients:

## Weeks worked

```{r}
model_dd_weeks_worked <- lm(wkswork ~ storm + everevac_bin + storm * everevac_bin, 
                            data = katrina)
model_dd_weeks_worked_full <- lm(wkswork ~ storm + everevac_bin + storm * everevac_bin +
                                   age + sex + black + hsgrad + someco + ba + postgrad, 
                            data = katrina)

tidy(model_dd_weeks_worked)
tidy(model_dd_weeks_worked_full)
```

```{r}
wks_worked_group_avgs <- katrina %>% 
  group_by(storm, everevac_bin) %>% 
  summarize(avg = mean(wkswork))
wks_worked_group_avgs
```

```{r}
# We assume that the evacuated group would have changed at the same rate as the non-evacuated group after the storm
# Not evacuated_after - Not evacuated_before
hypothetical_change <- 38.95388 - 38.23799

ggplot(wks_worked_group_avgs, aes(x = storm, y = avg, color = everevac_bin)) +
  geom_point() + 
  geom_line(aes(group = everevac_bin)) +
  geom_segment(aes(x = 1, y = 36.06564, xend = 2, 
                   yend = (36.06564 + hypothetical_change)), 
               linetype = "dotted", color = "grey10")
```

## Earnings

```{r}
# Models for earnings
```


## Unemployment income

```{r}
# Models for unempinc
```

